---
title: "Intro_to_R2AB"
output: html_document
date: "2024-03-05"
---
# Goals
* Describe and utilize built-in functions in R.
* Modify default behavior of a built-in function using arguments.
* Demonstrate how to create user-defined functions in R, including for loops etc.

## Introduction
Often when we use the R programming language it is for Data Analysis work, that is: managing large data sets, statistical analysis, and visualization of data. This will be how we use R in this short course, too. However, we are going to begin the course by spending time on using it as a programming language. This will lay a solid foundation for the data analysis and manipulation that we will learn to do over the remainder of the course (and the second half of the course, Intro to R II).

One of the major ideas in programming is that it allows you to break down a complex problem into more straightforward and tractable set of sub-problems. One of the tools available to us to do this is the mighty FOR loop (and, of course, the \*apply family which consists of: sapply, tapply, vapply, lapply, mapply - we saw this, briefly, is Module1B). The other tool is encapsulating repeated processes. We can do this by **defining our own functions**, instead of relying only on built-in functions. These are two new ideas that we will investigate in this Workbook (along the way, we'll look at anonymous functions, higher order functions and ellipsis and will use the built in package SWIRL() to help us learn this. 


#1. Built-in Functions: 
  * we can create our own bespoke function, but there are many functions that are part of R already (we can also access functions in packages)
  * you can accomplish a lot by changing the default arguments of built-in functions.
Let's examine a simple case: c(). This built in function takes a flexible number and type of arguments (can be strings, or numbers etc) and it works in slightly different ways depending on the order of arguments provided (you can add to the beginning or the end of an existing list):
```{r}
dogs<-c("Chihuhauas", "Golden Retreivers","German Shephards", "Great Pyrenese","Great Dane")
#we want to add a few dog breeds to the end of the this list and we can do that by using the original list as the first arguments
dogs<-c(dogs,"Norwegian elkhounds")
# we can see that it has added elkhounds to the end of the list
dogs
# how do we add to the beginning of the list?
dogs<-c("Pomeranians",dogs)
dogs
```
The core functionality of R is as a statistics and data analysis language so it is not surprising that many of the built-in functions have to do with math and data grooming: 
```{r}


```
#2. General Function Information: 
* Functions usually take in some sort of data structure (that is a value, a vector, a dataframe etc.), process/manipulate it, and return a result. The active part of the function is located between squiggly/curly braces. 

* Syntax: 
```
FunctionName(arguments){
    manipulation and processing of data
    return value
}
#You then want to call the function (which is how we have used all the built in functions that we 
#have seen so far in R such as mean(), median() etc)
function_name(value1, value2)
```
* there are functions that don't take arguments, but even these often have default values that are present (even if the user doesn't have to provide a value for them)

  
#3.User defined functions:
A. Functions

**The examples are mostly from SWIRL() - see section below on SWIRL() - and are vastly simplified. This means that if you get stuck, you can go back and work through the longer, more detailed examples in SWIRL().**

We can do nifty things with functions, like we can in other languages, such as setting default values in our functions. The function below demonstrates that you can use a default when creating a function and then, when you call that function, over-ride the default setting: 
```
increment <- function(number, by = 1){
   	 return(number + by)
}
    
increment(5)
```
any guesses what this would print to the screen? 
```
increment(5,4)
```
this should print out 9 since it has over ridden the default of 1 with the number 4
-----------------
R is particularly adept at statistical analysis of your data sets. However, what if you wanted to write your own statistics? Let's say we wanted to write a function that allowed us to calculate the **harmonic mean** of a data set (obviously we know that we could use the built-in mean() to find the arithmetic mean and could probably find a harmonic mean in a library but we want to write our own simple one).
This is an illustrative example of function syntax in R:
```{r}
# curly braces begin the action. If you are used to Python, this is a different syntax. Key words: function, return



```


B. For loop worked example:
 * repeatability, encapsulation and breaking down a problem is centrally important in programming and we can use for loops to achieve this. 
 * for loops have a particular syntax that needs to be followed, demonstrated in the cell below
    * we use curly braces in R: { defines the evaluated parts of the script. Basically, the 'action' needs to be between the {}
 * Remember that R IS WEIRD BECAUSE IT STARTS COUNTING AT 1 AND NOT 0. IT IS ALSO WEIRD BECAUSE IT IS **INCLUSIVE** AT THE END BRACKET!
 * There are a few built-in functions that we will use/see along the way
    * class(variable)
    * c()
    
```{r}
 for(i in 1:10){
  # this sentence will be repeated 10 times. You can modify the upper end of the 'slice' to give more times through or fewer time through the loop
  print("Remember that there is a syntax to defining your loops.RStudio makes it straightforward, though!")
}
# we need to have already defined a vector, x in order for the for loop to work. Let's define it here by using the c() function. 
# We don't know what the c() function does? How would we find out what it does? We could look up the functions online. 
# in fact, this is a commonly used function that we will see repeatedly! It COMBINES or CONCATENATES objects that you give it into a vector. 
x<-c(1,4,9,16,25,36,49,64,81,100)
print("*******")

# what happens if you want the for loop to repeat 11 times but give it a list of 10 as an argument. Most programming languages would throw an exception or give an error and it would halt the program until the error was resolved. R doesn't, though!
#------------------
for(i in 1:11){
  #this is how we can the value of each element in the vector: x[i]
 
  
  
}
#-----------
# instead of giving it a set number, we can base the number of loops on the length of the particular vector x that happens to be passed into the loop
# you will see this 'trick' a lot in other programming languages and it is 'best practices' for this type of loop.
for(i in 1:length(x)){

            
}
```

C. More worked examples of Functions!
#### A. Basics 
We will continue to create our own bespoke functions. Here is a function to draw a 'random number'. This function will call the built-in function runif() - which means that we don't have to define it since R already has done that for us- which randomly generates a uniform distribution of a particular size: 
```{r}



```
Here are a few other examples that are not list comprehensions (as the function in the chunk above is). 

R functions don't require an explicit return statements (many languages require return statements) but you should use one anyway so that you develop good programming habits. 
```{r}




```


##### B. Fancier than the basics: Higher order functions AKA \textbf{FUNCTIONS THAT CALL OTHER FUNCTIONS}
You can pass functions as arguments to other functions just like you can pass data to functions. Let's say you define the following functions:

```{r}




```
Before we work through some of the normal higher level functions, we should mention **anonymous functions**. We won't work through any list comprehensions since, in R, when a list comprehension is appropriate, we usually will use a *apply family function (such as lappy, tapply). Frankly, we will use the apply family for Map and sometimes Filter, too! However, since these higher order functions are present in so many other programming languages it is important to understand their logic.

Anonymous functions are useful functions that are not named (so they are not assigned to a name with the assignment operator). They are usually 'small' functions that will fit onto one line of code. They have a specific syntax that differs (but is based on) the generic function syntax. 

```{r}
#Say you have written the following simple function which takes a
# function and some data as arguments and applies the function to the data: 



```

R has elements of a functional programming language. You will also hear it called a "vector" language. As such, it relies on a number of higher order functions like Filter, Map, Reduce etc. Here are the higher order functions that are present in R (and in most other programming languages). Make sure that, once again, you spend some time trying to predict what the output of these variables will be:

1. Map
* applies the same formula to every element in a list
In R, we use the appropriate member of the apply family in the same way as Map. 
```{r}
hundred<-1:100
print(hundred)
class(hundred)




```

2. Filter
* selects elements in a vector that meet a the conditions of a criteria
```{r}
# Before hitting the run button on this cell, try to PREDICT what is 
# stored in the small.odd.nums variable
# The name is a pretty big hint but trying to predict the end result
# will help you understand what is happening in the code. 
# --------------------------------------------
# instead of defining the function and giving it a name, we can
# define 'anonymous' functions, which are nameless, and put them 
# into one line of a function


 

```

3. Reduce
* more limited than other higher order functions; takes two elements at a time and then manipulates them (like adding them together) and only then takes a third element and adds it to the running total. You often see Reduce used in biology when talking about the internal nodes (and their states) of phylogenies.  
```{r}
#example of reduce, which is more limited than other functions: 
#let's start with the painfully obvious and then refine the function definition:
my_sum <- function(x)
{
	if (length(x) == 1)
	{
		return(x[1])
	}
	if (length(x) == 2)
	{
		return(x[1]+ x[2])
	}
	if (length(x) == 3)
	{
		return(x[1] + x[2] + x[3])
	}
}
my_sum(c(1,4,9))
```

Let's take the above and make it more general by adding a loop; after all what happens if we have a something that is larger than 3? 
```{r}
my_sum2<-function(x)
{
	result<-0
 
	for (i in 1:length(x))
	{
		result<-result+x[i]
	}
 
	return(result)
}
#call the function just like we have called built in functions
my_sum2(c(1,4,9,25,36))
```
Let's compare this to the built in higher order reduce function: 
```{r}
# the name of the addition operation in R is `+`
my_sum3<- function (x) {Reduce(`+`, x)}
my_sum3(c(1,4,9,25,36))
# you can also multiply elements together: 
my.prod <- function (x) {Reduce(`*`, x)}
my.prod(c(1,4,9,25,36)) 
my.factorial<-function(n){Reduce(`*`, 1:n)}
my.factorial(c(7))
```

## ellipsis: 
```{r}
my_ellipsis <- function(...){
# unpack the ellpsis into a list:
args <- c(...)
sum(args)
}
my_ellipsis(1,2)
```

---------

### THIS SECTION IS ENTIRELY OPTIONAL IF YOU WANT TO EXPLORE SOME PROGRAMMING IN R CAPABILITIES. We won't cover this during classtime!
----------

# Learning to program in R with the package SWIRL()
We won't spend much time on programming in R, because we will mostly use R as a data analysis language, the typical way to use R. However, it can be used as a more general programming language, too. It normally isn't because - as a gross generalization - compared to other popular languages, it is not efficient with memory usage. In this section, I introduce you to SWIRL(). This is a package that you can work through yourself that will teach you to program in R. This is extra and isn't necessary, but I wanted to include it since I know that some individuals will want to spend time working through it! 

##### 1. R is mostly a functional (also sometimes referred to as a vector language) programming language.
This means that higher order functions - and list comprehensions - abound! What are higher order functions? Functions can take other functions as arguments. Higher order functions are functions that **take another function as an argument and then do something with it**. If you have familiarity with other programming languages, you will have seen higher order functions like filter, map, reduce and other less popular higher order functions such as find, position and negate. Note: We've already seen the \*apply family which is a lot like the higher order ** map ** function in other languages.

##### 2. The SWIRL() package is a fantastic entry point into R functionality.
swirl() will allow you to work through various programming concepts in an interactive environment. You will need to install the swirl package and load it. For the most part, the package itself will guide you through your options and is reasonably user-friendly. It is also an efficient, although a bit dry, way to quickly review concepts in the R language. It should not come as a surprise that programming languages often have to solve similar problems and do so in similar ways. 

**NOTE THAT YOU WILL NEED TO USE THE CONSOLE TO INTERACT WITH SWIRL(), NOT the markdown quadrant.**

SWIRL() includes the following chapters:

1: Basic Building Blocks      	  
2: Workspace and Files     
3: Sequences of Numbers           
4: Vectors                 
5: Missing Values            		  
6: Subsetting Vectors      
7: Matrices and Data Frames       
8: Logic                   
9: Functions                 		  
10: lapply and sapply       
11: vapply and tapply      	      
12: Looking at Data         
13: Simulation                		
14: Dates and Times         
15: Base Graphics 

You can always leave swirl() with the command bye(), and you can return to the main menu by typing main(). 
```{r}
#load the swirl library after installing the package so that knitter works!
library("swirl")
```
 
#### Review of key concepts using SWIRL()
Module1B focuses on the content from the SWIRL() library \textbf{Chapter 9: Functions}. However, the following chunk includes a summary of the other chapters of SWIRL() which are a basic review of material we have already seen (you should work through the chapters yourself if you are uncertain about your knowledge of the topics). 

Here is the simplified overview of the SWIRL() review chapters: 
```{r}
# Chapter 1 of Swirl(): Basic Building Blocks of R
# Especially important: the concatenate/combine function c()
Z<-c(1.1, 9,3.14)
class(Z)
Z*2+100 
list_1<-c(1,2,3)
list_2<-c(4,5,6)
list_3<-c(list_1,list_2)
class(list_3)
list_3
# You can skip Chapter 2: Workspace and files. However if you are interested in Linux/unix, you might enjoy working through the brief 
# introduction to file management

# Chapter3: Sequences and numbers
# Especially slicing. The ":" means that all numbers from lower to upper are included.
d<-1:20
d
# increase by 1 but since you are starting at 3.14, your output is a bit different.... 
e<-pi:10
e
# You can decrease as well
f<-15:1
f
# You can create a sequence of numbers by using the seq() function
# which also gives you the option of explicitly specifying an increment size.

g<-seq(1,20,by=1.5)
g
# rep() allows us to include an argument that specifies how many times
# something should be repeated. 
h<-rep(1:5,3)
h
# Chapter 4: Vectors
# Remember Boolean logic? Surprise - it is here again! R offers && versus & and || and | which behave differently on vectors. 
(3>5)|(4==4)
# paste()
my_char<-c("my", "name","is")
my_char
paste(my_char,collapse="")
#Chapter 5 addresses missing values: NA. 
#Chapter 6: subsetting vectors
# create a vector, X
X<-1:50
X
# Vector indexing: the first 10 elements
X[1:10]
# only include elements 3,6 and 8
X[c(3,6,8)]
# include all elements EXCEPT 2 and 41
X[c(-2,-41)]
X
# This is an example of named elements which means that instead of indexing element 1 or element 2, you can also assign them names like
# foo or orf or doo or any other ridiculous name that you can think of. In the example below, we have created a numeric list of two elements 
# where the first element is called 'foo' and the second element is "orf". Instead of calling an element by Y[1] as we could do if it was a regular
# vector, we can call the first element by its name, 'foo". The take home message is simply that we have more ways of indexing other than just 
# indexing with numbers.
Y<-c(foo=2,orf=10)
Y
names(Y)
Y["foo"]
# Chapter 7: Matrices and Data frames
# we create a vector called example_vector. Once that is created, we can run built in functions that exist for matrices and vectors. We can ask
# about specific attributes of the vector or matrix by running commands like class, attributes and length. 
example_vector<-1:20
class(example_vector)
# We can turn a vector into a matrix by passing the vector into the dim() function and specifying how many rows and column the final matrix should
# have. In this case, we have turned the vector into a matrix with 4 rows and 5 columns. 
dim(example_vector)<-c(4,5)
example_vector
class(example_vector)
attributes(example_vector)
length(example_vector)
# A bit like the c() function, there are two functions that bind lone vectors into a dataframe. rbind() combines rows and cbind() 
# combines columns. Remember the list_2 that we created above?We can bind two of them together to create a matrix.  
list_4<-rbind(list_2,list_2)
list_4
# Test what type of data they are now: 
class(list_4)
# Chapter 8: Boolean logic. Evaluate the following examples from swirl()
# && and || return a scalar TRUE/FALSE value whereas & and | return a vector
# Usually you would use &&, for instance, in a conditional loop to see if the specified condition has been met. 
FALSE || TRUE && FALSE
TRUE && FALSE || 9 >= 4 && 3 < 6
99.99 > 100 || 45 < 7.3 || 4 != 4.0
TRUE && 62 < 62 && 44 >= 44
```



.....and that's all, folks! I sincerely hope that you feel empowered to work in the R programming language!


